Challenging Fuel Cycle Modeling Assumptions: Facility and Time Step Discretization Effects

### 2015-10-22 paper.tex
no small caps cyclus
elliot's edits

% No more than three keywords, though each can be a phrase
\keywords{Nuclear Fuel Cycle, Simulation, \Cyclus}

@@

\myabstract{
Due to the diversity of fuel cycle simulator modeling assumptions, direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps one of the most complete published comparisons performed.  Despite
this, however, various results from the simulators were often significantly
different due to inconsistency in modeling decisions involving reprocessing
strategies, refueling behavior, reactor end-of-life handling, etc.  This work
attempts to identify and quantify the effects of selected modeling choices
that may sometimes be taken for granted in the fuel cycle modeling domain.
Four scenarios are compared using combinations of either fleet-based or
individual reactors with either monthly or quarterly (3-month) time steps.
The Cyclus fuel cycle simulator's plug-in facility capability along with its
market-like dynamic material routing allow it to be used as a level playing
field for comparing the scenarios.  When under supply-constraint pressure, the
four cases exhibit noticeably different behavior.  Fleet-based modeling is more
efficient in supply-constrained environments at the expense of losing insight
on issues such as realistically suboptimal fuel distribution and challenges in
reactor refueling cycle staggering.  Finer-grained time steps also enable more
efficient material use in supply-constrained environments.  Large simulations
with fleet-based reactors run much more quickly than their individual reactor
counterparts.
}

@@

The diversity of assumptions embedded within various fuel cycle simulators pose challenges for direct
comparison and benchmarking.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study
\cite{oecd2012benchmark} that is perhaps one of the most complete published
comparisons performed.  Despite this, however, various results from the
different simulators were often significantly different due to inconsistent
modeling decisions involving reprocessing strategies, refueling behavior,
reactor end-of-life handling, etc.  This work attempts to identify and
quantify the effects of selected modeling choices relative to facility modeling
discretization (i.e. individual facilities vs. aggregate fleets) and time step
duration.  A better understanding of how these affect fuel cycle dynamics can
allow for making more deliberate decisions with respect to trade-offs such as
computational investment vs. realism.

@@

Traditionally, many fuel cycle simulators have used a system dynamics approach
\cite{forrester_industrial_1961} to modeling. These simulators model several
"stocks" representing isotopes or isotope groups that have levels that change
over time due to "flows" between them.  The flows are determined by equations
that are a function of the state of the system (i.e. the system's stocks).  A
constructed system is then solved in discrete or continuous time depending on
the limitations of the software and constraints imposed by the model.  In
system dynamics based simulators, the number of stocks and potential flows between
them is generally a static property of the modeled system, making it somewhat
difficult to swap different reactor models in and out of the same scenario
without wiring them into the simulator manually.

@@

Many fuel cycle simulators step through time in discrete steps.  The duration
of a single time step can have significant impact on the dynamics of a
simulation.  The time step duration can affect the magnitude of individual
material transfers between facilities and correspondingly affects the noise
level in facility material inventories and flows.  It can also act as a
minimum bound for outage durations and is a minimum bound in general for state
changes within the entire system. The effects of the time step duration are
also investigated together with fleet vs facility modeling.

@@

\subsection{\Cyclus}

@@

As a Cyclus simulation walks through each time step, facilities engage in a
market-like dynamic resource exchange (DRE).  They broadcast requests for
material and potential supplying facilities respond with bids. The requesting facilities then
express preferences between all their bids, and the DRE then generates a network
flow problem that is optimized to resolve which transactions actually occur.
The plug-in architecture combined with the DRE creates a powerful and flexible
environment for the natural comparison of different modeling paradigms.
Cyclus' module plug-in based architecture in concert with its DRE has the ability to
model fleet-based facilities.  Additionally, however, Cyclus can model
facilities individually with discrete flows enabling the investigation of
interesting real-world effects such as competition, reactor cycle staggering,
and individual reactor outages.

@@

    \item Refueling occurs continuously with as much fresh fuel as is
        available up to the total amount to fill all reactor cores in the
        fleet.  No extra fuel is kept on-hand - it is all ordered
        just-in-time.

@@

    \item When a reactor in the fleet is retired, a full core of material is
        discharged - even if there is some fraction of the fleet with unfueled
        cores.

@@

The scenario starts with 100 LWRs and follows an exponential curve with a 1\%
annual growth rate for 200 years.  The decommissioning of the initial 100
reactors is staggered over years 15 to 55.  Fresh fuel for LWRs is provided directly from enrichment with infinite supply and throughput. In year 35, fast reactors become available for
deployment and no more thermal reactors are built.  LWR spent fuel separations
begins in year 15 with 2000 MTHM/yr capacity and increases to 3000 MTHM/yr in
year 25.  Fast reactor spent fuel separations and fuel fabrication have
infinite capacity.  All spent fuel is stored/cooled for 7 years before it becomes
available for separation: 5 years for cooling and an additional 2 years to approximate non-instantaneous separations and fabrication times. Figure \ref{fig:flow} shows the material flow
relationships between the facilities.  Compositions used for the simulations
were simple containing $^{238}U$, $^{235}U$, $^{239}Pu$, $^{241}Am$,
and small quantities of a few lanthanides.  Full composition details can be
found in the Cyclus input files used \cite{Carlsen2015}.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.47\textwidth]{exp2/flow.eps}
    \caption[Material flow paths]{Potential material flow paths between facility types.}
    %% consider a Cyclist-derived visualization here?
    \label{fig:flow}
\end{figure}

%% is it worth waiting until the 1.4 release?  Citing an ambiguous pre-release makes
%% this a little harder to reproduce.
A pre-release of Cyclus version 1.4 was used for this analysis, with the
standard archetypes provided with Cyclus in the included Cycamore
library - also a pre-release of version 1.4.  A custom storage archetype was
developed because Cycamore did not yet provide one.  The standard Cycamore reactor archetype
was used for both LWR and SFR reactor types in cases MI and QI.  Correspondingly
the custom fleet-based reactor archetype described earlier was used for both reactor
types in cases MF and QF.  Additional configuration for the reactor types is
shown in Table \ref{tab:reactors} (cycle length includes the refueling outage).

@@

\begin{table*}
    \centering
    \captionsetup{justification=centering}
    \caption{Selected Reactor Parameters by Case}
    \begin{tabular}{ |r | c c | c c | }
        \hline                       
                                          & \multicolumn{2}{c |}{LWR}       & \multicolumn{2}{c |}{SFR} \\
                                          & Cases MI,QI & Cases MF,QF (fleet) & Cases MI,QI & Cases MF,QF (fleet)  \\
        \hline                       
        Cycle length (months)             & 15        & 18                & 12        & 15 \\
        Refueling outage (months)         & 3         & 0                 & 3         & 0 \\
        Batch size ($\text{kg} \cdot \text{HM}$) & 29,565    & 29,565            & 8,025     & 8,025 \\
        Power capacity (MWe)              & 1,080     & 900               & 450       & 360 \\
        \hline                       
    \end{tabular}
    \label{tab:reactor-detail}
\end{table*}

In each of the four cases, the initial 100 LWRs are modified to have a zero
length refueling outage with an explicit 900 MWe net power output capacity.  Because the
individual reactor model doesn't have any way to start a reactor mid-cycle,
this was done to avoid having all initial LWRs refuel at the same time.  The
parameters for each case were also selected to keep capacity factors for
individual reactors very roughly around 90\%, although they are actually 
closer to 80\% due to integer variable constraints on the case MI and QI
reactor configuration.

@@

There are various low level phenomena that have potential to affect the larger
outcome of simulations with respect to facility discretization and time step
modeling decisions. Several effects arising from these different modeling
choices are present in many fuel cycle simulators to varying degrees.  Some of
these effects are described in detail below.

One difference between the fleet and individual facility modeling is the
\emph{quantized shutdown effect}.  If a fleet-based reactor modeling 3-batch
cores is short one half batch of fuel, it will produce $\frac{2.5}{3}\cdot P_r$
of power for that time step where $P_r$ is the power capacity of a single
reactor of the fleet (\emph{e.g.}~1000~MWe).  If an individually modeled reactor is
short one half batch of fuel, then it will produce no power ($0 \cdot P_r$) that
time step.  Time step duration can amplify this effect.  For both fleet and
individual reactor modeling, power goes offline in full time step increments.
Consider an example of a 3-batch reactor with a fuel shortage of $\frac{1}{2}$ batch of
fuel that would be resolve in 1 month.  Depending on the modeling choice, each reactor
would produce the following amounts of energy over a 3 month period:

\begin{itemize}
  \item case \emph{QF}: $\frac{2.5}{3} \cdot 3 \text{ mo} \cdot P_r = 2.5P_r$
  \item case \emph{MF}: $[\frac{2.5}{3} \cdot 1 \text{ mo} + 1 \cdot 2
    \text{mo}] \cdot P_r = 2.83P_r$
  \item case \emph{QI}: $0 \cdot 3 \text{ mo} \cdot P_r = 0$
  \item case \emph{MI}: $[0\cdot 1 \text{ mo} + 1 \cdot 2\text{mo}] \cdot P_r = 2 P_r$
\end{itemize}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1.0\columnwidth]{exp2/sync-cycle.eps}
    \caption[The cycle staggering effect]{
        In this scenario, reactors are deployed annually with a refueling cycle
        length (including outage) of 18 months.  So all reactors deployed
        every 3rd year refuel together.  All initial 100 LWRs start out with
        their cycles synced as well.
    }
    \label{fig:sync-cycle}
\end{figure}

Another difference between fleet-based and individual facility modeling
involves fuel sharing.  Power production by a group of individual reactors for
a fixed amount of fuel shortage may vary depending on how available fuel is
distributed.  This is illustrated in Figure \ref{fig:fuel-sharing}.  A system
being short 3 batches might mean that one reactor has no batches and isn't
operating or that 3 reactors are each missing a single batch.  This is
referred to as the \emph{fuel sharing effect}.  Fleet reactors, by design,
have perfect fuel sharing - all fuel is distributed so that the minimum
possible amount of capacity is offline during shortages. The most correct
measure of fuel sharing inefficiency is the difference between the actual
generated power and the maximum amount of power that could have been generated
among all possible fuel distribution alternatives.  This is difficult to
actually compute. One possible approximation is to count the number of fresh
fuel batches distributed to reactors that ended up not having enough fuel to
operate (i.e. a full core) that time step.  This approximation assumes that
every "wasted" batch of fuel, could have been given to a reactor that only
needed one batch to operate.  Equation \ref{eq:badshare} describes such an
approximation.

@@

where $R_t$ is the set of all reactors, $H$ is the Heaviside function,
$\tau(t,r)$ is the beginning of the most recent refueling outage of reactor $r$ before $t$,
$\Delta\; t_{out}$ is the normal refueling outage duration, $F_{prev}(t,r)$ is
the start of the most recent refueling outage for reactor $r$ before or on
time $t$, $O(t,r)$ is 1 if reactor $r$ is producing power at time $t$ and zero
otherwise, and $N_b(t,r)$ is the number of new fuel batches received by
reactor $r$ since time $t$.

@@

There are also performance differences between the two modeling paradigms.  A
single 200 year simulation with several hundred fleet-based reactors takes
about 2 seconds to run on a typical desktop machine where the same simulation
using individual reactor facilities takes about 20 seconds.  This order of
magnitude difference is an important trade-off to keep in mind when doing fuel
cycle analysis in general, particularly when considering optimization or
sensitivity studies where running thousands or millions of simulations may be
desirable.  The time step duration has less impact on performance because the
amount of work being done in the simulation is similar; in the monthly time
step cases most facilities (i.e. all reactors) only perform actions every
refueling cycle rather than every time step.

@@

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/power.eps}
    \caption[Generated power]{
        Generated power for all four cases.  A fuel shortage occurs from about
        year 100 to about year 140.  In cases MI and QI with individual reactor
        modeling, refueling cycles become much more synchronized during and
        after the fuel shortage resulting in large power fluctuations - an artifact of the
        \emph{cycle staggering effect}.
    }
    \label{fig:power}
\end{figure*}

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/power-rel.eps}
    \caption[Normalized power]{
        Generated power is normalized to the expected 1\% exponential growth
        curve for all four cases.  A fuel shortage occurs from about year 100
        to about year 140.  The \emph{cycle staggering effect} can be seen as
        excessive divergence above and below $1.0$ in cases MI and QI where
        the shortage increases cycle synchronization significantly. The
        \emph{quantized shutdown effect} can also be seen in the discrepancy
        between quarterly and monthly time steps for fleet reactor modeling
        (cases MF and QF) during the shortage; a longer time step causes more
        reactor outage than necessary. The consistent jumping up and down for
        the fleet reactor cases (even pre-shortage) is due to the mis-alignment of reactors going
        offline (multiples of 12 months) and the build period (21 months).
    }
    \label{fig:power-rel}
\end{figure*}

@@

where $R_t$ is the set of all reactors, $P_r$ is the power capacity of reactor
$r$, $H$ is the Heaviside function, $\tau(t,r)$ is the beginning of the most
recent refueling outage of reactor $r$ before $t$, $\Delta\; t_{out}$ is the
normal refueling outage duration, and $O(t,r)$ is 1 if reactor $r$ is
producing power at time $t$ and zero otherwise.

@@

Quantifying imperfect fuel sharing is a bit difficult, but Figure
\ref{fig:badshare} provides one way to see the \emph{fuel sharing effect}
showing wasted batch-months computed using Equation \ref{eq:badshare}.  Fleet
reactor modeling, by design, exhibits perfect fuel sharing (zero
inefficiency) and so is not included in this figure.  Every batch that is
given to a reactor that ends up not being able to start its cycle at the
scheduled time (due to fuel shortage) adds one to this cumulative for each
time step the cycle start is delayed.

@@

One useful over-estimate of the fuel sharing inefficiency is to assume every
one of the "wasted" batches could have been given to a reactor enabling it to
operate.  Multiplying each of these cumulative batches by the fast reactor power capacity
(i.e. 450 MWe) provides a way to compare fuel sharing inefficiency with the
overall fuel outage.  This results in a cumulative energy deficit caused by
poor fuel sharing of roughly 2,000 $\text{GWe} \cdot \text{months}$ and 6,600
$\text{GWe} \cdot \text{months}$ for cases MI and QI respectively.  These
approximations account for a significant portion of the difference between individual
and fleet cases in Figure \ref{fig:unfueled} suggesting that most of the
discrepancy in cumulative outage between fleet and individual modeling is
caused by inefficient fuel sharing in the individual reactor cases.

@@

Having multiple assemblies per batch (item 1 above) will further degrade fuel
sharing efficiency.  Even reactors that are short only one batch could
potentially receive some fuel and still be unable to operate.  In order to
maintain optimal sharing, fresh fuel now needs to be sent in all-or-nothing
multi-assembly quanta with higher preference to reactors that need fewer
assemblies.  Spare fresh fuel inventory (item 2 above) will have the global
effect of increasing total shortage severity; on average more batches of fresh
fuel will be sitting unused.  However, it can potentially reduce the frequency
of outages for individual reactors in some cases.

@@

The large differences seen between monthly and quarterly time step cases in
Figures \ref{fig:unfueled} and \ref{fig:badshare} are primarily caused by the
\emph{inventory drawdown effect}.  Figure \ref{fig:puinv} helps to visualize
this effect.  The black dots in the figure represent impulse flows of
available Pu from fuel fabrication and into fast reactors.  Blue lines represent
the flows into Pu inventory available for making fuel. The in-flows are not
shown for the individual reactor cases because they are somewhat messy and
obscure other information in the plots. The other colored curves show Pu inventory
available for fabricating fast reactor fuel.  When the inventory curves meet and
force down the out-flows, fuel shortages occur.  While this is most clear for
cases MF and QF, the individual reactor scenarios (cases MI and QI) also have many
points during the shortage where the outflow point lies exactly on top of the
inventory point - indicating that all available inventory is being
transferred.

@@

Throughout most of the shortage, the aggregate outflow of Pu from separated
inventory is roughly the same in cases MF and QF.  However, due to the longer
time steps in case QF, the shortage starts slightly sooner because Pu inventory
must be maintained at a higher limit as described above.  The small difference
in outage start time has feed-back effects.  Not only are extra reactors
offline, but those extra reactors are also discharging less aggregate spent
fuel for recycling.  This increases the discrepancy between the two
cases. Eventually the shortage in case MF ends slightly before it does in case
QF (shown most clearly in Figure \ref{fig:puinv-compare-zoom}). This feedback effect occurs because the reactors have a
conversion ratio greater than 1.0.

The horizontal striations visible in the monthly flows for case MI in Figure
\ref{fig:puinv} are a characteristic of the synchronization of reactor
refueling caused by the shortage.  A close-up view of these can be seen in
Figure \ref{fig:puinv-MI}. They begin at the end of the fuel shortage window
when many reactors come online together in groups and are resonances that the
inventory level jumps between.  These resonances gradually disappear as the
simulation progresses farther in time beyond the shortage and refueling cycles
become more staggered.

@@

The reactors in all the scenarios request fresh fuel only when they need it.
While this is an unrealistic behavior for real-world facilities, it is closer
to a globally optimum fuel management strategy.  Keeping on-hand fresh fuel
batches would also have the same effect of increasing idling plutonium
quantities and would make the fuel shortage worse overall.

>>>>

% No more than three keywords, though each can be a phrase
\keywords{Nuclear Fuel Cycle, Simulation, Cyclus}

@@

\myabstract{
Due to the diversity of fuel cycle simulator modeling assumptions, direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps the most complete published comparisons performed.  Despite this,
various results from the simulators were often significantly different due to
inconsistency in modeling decisions involving reprocessing strategies,
refueling behavior, reactor end-of-life handling, etc.  This work identifies
and quantifies the effects of selected modeling choices that may sometimes be
taken for granted in the fuel cycle modeling domain.  Four scenarios are
compared using combinations of either fleet-based or individual reactors with
either monthly or quarterly (3-month) time steps.  The scenarios approximate a
transition from the current once-through light water reactor (LWR) fleet to a
full sodium fast reactor (SFR) fuel cycle The Cyclus fuel cycle simulator's
plug-in facility capability along with its market-like dynamic material
routing allow it to be used as a level playing field for comparing the
scenarios.  When under supply-constraint pressure, the four cases exhibit
noticeably different behavior.  Fleet-based modeling is more efficient in
supply-constrained environments at the expense of losing insight on issues
such as realistically suboptimal fuel distribution and challenges in reactor
refueling cycle staggering.  Finer-grained time steps also enable more
efficient material use in supply-constrained environments.  Large simulations
with fleet-based reactors run much more quickly than their individual reactor
counterparts.  Gaining a better understanding of how these and other modeling
choices affect fuel cycle dynamics will enable making more deliberate
decisions with respect to trade-offs such as computational investment vs.
realism.
}

@@

The diversity of assumptions embedded within various fuel cycle simulators pose challenges for direct
comparison and benchmarking.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study
\cite{oecd2012benchmark} that is perhaps one of the most complete published
comparisons performed.  Despite this, however, various results from the
different simulators were often significantly different due to inconsistent
modeling decisions involving reprocessing strategies, refueling behavior,
reactor end-of-life handling, etc.  This work attempts to identify and
quantify the effects of selected modeling choices relative to facility modeling
discretization (i.e. individual facilities vs. aggregate fleets) and time step
duration.  A better understanding of how these affect fuel cycle dynamics will
allow for making more deliberate decisions with respect to trade-offs such as
computational investment vs. realism.

@@

Traditionally, many fuel cycle simulators have used a system dynamics approach
\cite{forrester_industrial_1961} to modeling. These simulators model several
\emph{stocks} representing isotopes or isotope groups that have quantities that change
over time due to \emph{flows} between them.  The flows are determined by equations
that are a function of the state of the system (i.e. the system's stocks).  A
constructed system is then solved in discrete or continuous time depending on
the limitations of the software and constraints imposed by the model.  In
system dynamics based simulators, the number of stocks and potential flows between
them is generally a static property of the modeled system, making it somewhat
difficult to swap different reactor models in and out of the same scenario
without wiring them into the simulator manually.

@@

Many fuel cycle simulators step through time in discrete steps.  The duration
of a single time step can have significant impact on the dynamics of a
simulation.  The time step duration can affect the magnitude of individual
material transfers between facilities and correspondingly affects the noise
level in facility material inventories and flows.  It can also act as a
minimum bound for outage durations and is a minimum bound in general for state
changes within the entire system. The effects of the time step duration are
also investigated together with fleet vs. facility modeling.

@@

\subsection{Cyclus}

@@

As a Cyclus simulation walks through each time step, facilities engage in a
market-like dynamic resource exchange (DRE).  They broadcast requests for
material and potential supplying facilities respond with bids. The requesting facilities then
express preferences between all their bids, and the DRE then generates a network
flow problem that is optimized to resolve which transactions actually occur.
The plug-in architecture combined with the DRE creates a powerful and flexible
environment for the natural comparison of different modeling paradigms.
Cyclus' module plug-in based architecture in concert with its DRE has the ability to
model fleet-based facilities.  Additionally, Cyclus can model
facilities individually with discrete flows enabling the investigation of
real-world effects such as competition, reactor cycle staggering,
and individual reactor outages.

@@

    \item Refueling occurs continuously with as much fresh fuel as is
        available up to the total amount to fill all reactor cores in the
        fleet.  No extra fuel is kept on-hand --- it is all ordered
        just-in-time.

@@

    \item When a reactor in the fleet is retired, a full core of material is
        discharged --- even if there is some fraction of the fleet with unfueled
        cores.

@@

The scenario starts with 100 LWRs and follows an exponential curve with a 1\%
annual growth rate for 200 years.  The decommissioning of the initial 100
reactors is staggered over years 15 to 55.  Fresh fuel for LWRs is provided directly from enrichment with infinite supply and throughput. In year 35, fast reactors become available for
deployment and no more thermal reactors are built.  LWR spent fuel separations
begins in year 15 with 2,000 MTHM/yr capacity and increases to 3,000 MTHM/yr in
year 25.  Fast reactor spent fuel separations and fuel fabrication have
infinite capacity.  All spent fuel is stored/cooled for 7 years before it becomes
available for separation: 5 years for cooling and an additional 2 years to approximate non-instantaneous separations and fabrication times. Figure \ref{fig:flow} shows the material flow
relationships between the facilities.  Compositions used for the simulations
were simple containing $^{238}U$, $^{235}U$, $^{239}Pu$, $^{241}Am$,
and small quantities of a few lanthanides.  Full composition details can be
found in the Cyclus input files used \cite{Carlsen2015}.

\begin{figure}[!h]
    \centering
    \includegraphics[keepaspectratio=true,width=0.45\textwidth,height=.9\textheight]{exp2/flow.eps}
    \caption[Material flow paths]{Potential material flow paths between facility types.}
    %% consider a Cyclist-derived visualization here?
    \label{fig:flow}
\end{figure}

%% is it worth waiting until the 1.4 release?  Citing an ambiguous pre-release makes
%% this a little harder to reproduce.
A pre-release of Cyclus version 1.4 was used for this analysis.  A standard library of facility plug-ins, Cycamore, is provided with Cyclus.
Like Cyclus, a pre-release of Cycamore version 1.4 was used.  A custom storage facility archetype was
developed because Cycamore did not yet provide one.  Code for the fleet reactor and
storage archetypes is publicly available and can be downloaded \cite{Carlsen2015}. The standard Cycamore reactor archetype
was used for both LWR and SFR reactor types in cases MI and QI.  Correspondingly
the custom fleet-based reactor archetype described earlier was used for both reactor
types in cases MF and QF.  Additional configuration for the reactor types is
shown in Table \ref{tab:reactors} (cycle length includes the refueling outage).

@@

\begin{table*}
    \centering
    \captionsetup{justification=centering}
    \caption{Selected Reactor Parameters by Case}
    \begin{tabular}{ |r | c c | c c | }
        \hline                       
                                          & \multicolumn{2}{c |}{LWR}       & \multicolumn{2}{c |}{SFR} \\
                                          & Cases MI, QI & Cases MF, QF & Cases MI, QI & Cases MF, QF  \\
        \hline                       
        Cycle length (months)             & 15        & 18                & 12        & 15 \\
        Refueling outage (months)         & 3         & 0                 & 3         & 0 \\
        Batch size ($\text{kg} \cdot \text{HM}$) & 29,565    & 29,565            & 8,025     & 8,025 \\
        Power capacity (MWe)              & 1,080     & 900               & 450       & 360 \\
        \hline                       
    \end{tabular}
    \label{tab:reactor-detail}
\end{table*}

In each of the four cases, the initial 100 LWRs are modified to have a zero-length
refueling outage with an explicit 900 MWe net power output capacity.  Because the
individual reactor model does not have any way to start a reactor mid-cycle,
this was done to avoid having all initial LWRs refuel at the same time.  The
parameters for each case were also selected to keep capacity factors for
individual reactors very roughly around 90\%, although they are actually 
closer to 80\% due to integer variable constraints on the case MI and QI
reactor configuration.

@@

There are various low-level phenomena that have potential to affect the larger
outcome of simulations with respect to facility discretization and time step
modeling decisions. Several effects arising from these different modeling
choices are present in many fuel cycle simulators to varying degrees.  Some of
these effects are described in detail below.

One difference between the fleet and individual facility modeling is the
\emph{quantized shutdown effect}.  If a fleet-based reactor modeling 3-batch
cores is short one half batch of fuel, it will produce $\frac{2.5}{3}\cdot P_r$
of power for that time step where $P_r$ is the power capacity of a single
reactor of the fleet (e.g.~1000~MWe).  If an individually modeled reactor is
short one half batch of fuel, then it will produce no power ($0 \cdot P_r$) that
time step.  Time step duration can amplify this effect.  For both fleet and
individual reactor modeling, power goes offline in full time step increments.
Consider an example of a 3-batch reactor with a fuel shortage of $\frac{1}{2}$ batch of
fuel that would be resolved in 1 month.  Depending on the modeling choice, each reactor
would produce the following amounts of energy over a 3 month period:

\begin{itemize}
  \item case \emph{QF}: $\frac{2.5}{3} \cdot 3 \text{ mo} \cdot P_r = 2.5P_r$
  \item case \emph{MF}: $[\frac{2.5}{3} \cdot 1 \text{ mo} + 1 \cdot 2
    \text{mo}] \cdot P_r = 2.83P_r$
  \item case \emph{QI}: $0 \cdot 3 \text{ mo} \cdot P_r = 0$
  \item case \emph{MI}: $(0\cdot 1 \text{ mo} + 1 \cdot 2\text{ mo}) \cdot P_r = 2 P_r$
\end{itemize}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1.0\columnwidth]{exp2/sync-cycle.eps}
    \caption[The cycle staggering effect]{
        In this scenario, reactors are deployed annually with a refueling cycle
        length (including outage) of 18 months.  As a result, all reactors deployed
        every 3rd year refuel together.  All initial 100 LWRs start out with
        their cycles synced as well.
    }
    \label{fig:sync-cycle}
\end{figure}

Another difference between fleet-based and individual facility modeling
involves fuel sharing.  Power production by a group of individual reactors for
a fixed amount of fuel shortage may vary depending on how available fuel is
distributed.  This is illustrated in Figure \ref{fig:fuel-sharing}.  A system
being short 3 batches might mean that one reactor has no batches and is not
operating or that 3 reactors are each missing a single batch.  This is
referred to as the \emph{fuel sharing effect}.  Fleet reactors, by design,
have perfect fuel sharing --- all fuel is distributed so that the minimum
possible amount of capacity is offline during shortages. The most correct
measure of fuel sharing inefficiency is the difference between the actual
generated power and the maximum amount of power that could have been generated
among all possible fuel distribution alternatives.  This is difficult to
actually compute. One possible approximation is to count the number of fresh
fuel batches distributed to reactors that ended up not having enough fuel to
operate (i.e. a full core) that time step.  This approximation assumes that
every "wasted" batch of fuel, could have been given to a reactor that only
needed one batch to operate.  Equation \ref{eq:badshare} describes such an
approximation.

@@

\noindent
where $R_t$ is the set of all reactors, $H$ is the Heaviside function,
$\tau(t,r)$ is the beginning of the most recent refueling outage of reactor $r$ before $t$,
$\Delta\; t_{out}$ is the normal refueling outage duration, $F_{prev}(t,r)$ is
the start of the most recent refueling outage for reactor $r$ before or on
time $t$, $O(t,r)$ is 1 if reactor $r$ is producing power at time $t$ and zero
otherwise, and $N_b(t,r)$ is the number of new fuel batches received by
reactor $r$ since time $t$.

@@

There are also performance differences between the two modeling paradigms.  A
single 200 year simulation with several hundred fleet-based reactors takes
about 2 seconds to run on a computer with an Intel Core(TM) i7-4770 3.40GHz CPU where the same simulation
using individual reactor facilities takes about 20 seconds.  This order of
magnitude difference is an important trade-off to keep in mind when doing fuel
cycle analysis in general, particularly when considering optimization or
sensitivity studies where running thousands or millions of simulations may be
desirable.  The time step duration has less impact on performance because the
amount of work being done in the simulation is similar; in the monthly time
step cases most facilities (i.e. all reactors) only perform actions every
refueling cycle rather than every time step.

@@

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/power.eps}
    \caption[Generated power]{
        Generated power for all four cases.  A fuel shortage occurs from about
        year 100 to about year 140.  In cases MI and QI with individual reactor
        modeling, refueling cycles become much more synchronized during and
        after the fuel shortage resulting in large power fluctuations --- an artifact of the
        \emph{cycle staggering effect}.
    }
    \label{fig:power}
\end{figure*}

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/power-rel.eps}
    \caption[Normalized power]{
        Generated power is normalized to the expected 1\% exponential growth
        curve for all four cases.  A fuel shortage occurs from about year 100
        to about year 140.  The \emph{cycle staggering effect} can be seen as
        excessive divergence above and below $1.0$ in cases MI and QI where
        the shortage increases cycle synchronization significantly. The
        \emph{quantized shutdown effect} can also be seen in the discrepancy
        between quarterly and monthly time steps for fleet reactor modeling
        (cases MF and QF) during the shortage; a longer time step causes more
        reactor outage than necessary. The consistent jumping up and down for
        the fleet reactor cases (even pre-shortage) is due to the misalignment of reactors going
        offline (multiples of 12 months) and the build period (21 months).
    }
    \label{fig:power-rel}
\end{figure*}

@@

\noindent
where $R_t$ is the set of all reactors, $P_r$ is the power capacity of reactor
$r$, $H$ is the Heaviside function, $\tau(t,r)$ is the beginning of the most
recent refueling outage of reactor $r$ before $t$, $\Delta\; t_{out}$ is the
normal refueling outage duration, and $O(t,r)$ is 1 if reactor $r$ is
producing power at time $t$ and zero otherwise.

@@

Quantifying imperfect fuel sharing is a bit difficult, but Figure
\ref{fig:badshare} provides one way to see the \emph{fuel sharing effect}
showing wasted batch-months computed using Equation \ref{eq:badshare}.  Fleet
reactor modeling, by design, exhibits perfect fuel sharing (zero
inefficiency) and so is not included in this figure.  Every batch that is
given to a reactor that ends up not being able to start its cycle at the
scheduled time (due to fuel shortage) adds one to this cumulative total for each
time step the cycle start is delayed.

@@

One useful overestimate of the fuel sharing inefficiency is to assume every
one of the "wasted" batches could have been given to a reactor enabling it to
operate.  Multiplying each of these cumulative batches by the fast reactor power capacity
(i.e. 450 MWe) provides a way to compare fuel sharing inefficiency with the
overall fuel outage.  This results in a cumulative energy deficit caused by
poor fuel sharing of roughly 2,000 $\text{GWe} \cdot \text{months}$ and 6,600
$\text{GWe} \cdot \text{months}$ for cases MI and QI respectively.  These
approximations account for a significant portion of the difference between individual
and fleet cases in Figure \ref{fig:unfueled} suggesting that most of the
discrepancy in cumulative outage between fleet and individual modeling is
caused by inefficient fuel sharing in the individual reactor cases.

@@

Having multiple assemblies per batch (item 1 above) will further degrade fuel
sharing efficiency.  Even reactors that are short only one batch could
potentially receive some fuel and still be unable to operate.  In order to
maintain optimal sharing, fresh fuel now needs to be sent in all-or-nothing
multi-assembly quanta with higher preference to reactors that need fewer
assemblies.  Spare fresh fuel inventory (item 2 above) will have the global
effect of increasing total shortage severity; on average more batches of fresh
fuel will be idling unused.  However, it can potentially reduce the frequency
of outages for individual reactors in some cases.

@@

The large differences seen between monthly and quarterly time step cases in
Figures \ref{fig:unfueled} and \ref{fig:badshare} are primarily caused by the
\emph{inventory drawdown effect}.  Figure \ref{fig:puinv} helps to visualize
this effect.  The black dots in the figure represent impulse flows of
available Pu from fuel fabrication and into fast reactors.  Blue lines represent
the flows into Pu inventory available for making fuel. The in-flows are not
shown for the individual reactor cases because they are somewhat messy and
obscure other information in the plots. The other colored curves show Pu inventory
available for fabricating fast reactor fuel.  When the inventory curves meet and
force down the out-flows, fuel shortages occur.  While this is most clear for
cases MF and QF, the individual reactor scenarios (cases MI and QI) also have many
points during the shortage where the outflow point lies exactly on top of the
inventory point --- indicating that all available inventory is being
transferred.

@@

Throughout most of the shortage, the aggregate outflow of Pu from separated
inventory is roughly the same in cases MF and QF.  However, due to the longer
time steps in case QF, the shortage starts slightly sooner because Pu inventory
must be maintained at a higher limit as described above.  The small difference
in outage start time has feedback effects.  Not only are extra reactors
offline, but those extra reactors are also discharging less aggregate spent
fuel for recycling.  This increases the discrepancy between the two
cases. Eventually the shortage in case MF ends slightly before it does in case
QF (shown most clearly in Figure \ref{fig:puinv-compare-zoom}). This feedback effect occurs because the reactors have a
conversion ratio greater than 1.0.

The horizontal striations visible in the monthly flows for case MI in Figure
\ref{fig:puinv} are a characteristic of the synchronization of reactor
refueling caused by the shortage.  A close-up view of these can be seen in
Figure \ref{fig:puinv-MI}. They begin at the end of the fuel shortage window
when many reactors come online together in groups and are resonances that the
inventory level jumps between.  This artifact is associated with poor cycle staggering
like that shown earlier in Figure \ref{fig:sync-cycle}.
These resonances gradually disappear as the
simulation progresses farther in time beyond the shortage and refueling cycles
become more staggered.

@@

The reactors in all the scenarios request fresh fuel only when they need it.
While this is an unrealistic behavior for real-world facilities, it is closer
to a globally optimum fuel management strategy.  Keeping on-hand fresh fuel
batches would also have the same effect of increasing idling Pu
quantities and would make the fuel shortage worse overall.
